---
title: "Data Visualisation Project"
author: "Seline Uran"
output: html_document
---
```{r = diagram1, echo=FALSE, message=FALSE}
#To begin load the following libraries
library(ggplot2)
library(readxl)
library(RColorBrewer)
library(scales)
library(dplyr)
library(tidyr)
library(tidyverse)
library(gridExtra)
library(RColorBrewer)
```

<span style="text-decoration:underline">Research Questions</span>

These graphs present government survey data on the number of rough sleepers in England to consider the following questions; 

Has the numer of rough sleepers in England increased over the past decade? 

Is the numer of sleepers observed equally across England in one year? 

<span style="text-decoration:underline">Data Origins</span>

XXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXX

<span style="text-decoration:underline">Data Preparation</span>

After having removed all sheets from the excel document except for the sheet that contained the total number of rough sleepers I began data preparation. 

```{r}
#importing the raw data
rawdata <- read_excel("~/myfinalproject/processed/rs_statistics.xlsx")
```

The raw data detailed the number of rough sleepers within England and individual counts for each region of England and each Local Authority within each region. 

My aim was to visualise the number of rough sleepers in each region of England across the last 10 years. I selected these rows from the data set. Since I aimed to visualize the number of rough sleepers in each region individually I removed the rows containing summary counts for 'England', 'London' and the 'Rest of England'. Then, I removed data associated with each Local Authority, columns containing the ONS code of each Region and Local Authority. 

```{r}
#select the columns required
rs1<- rawdata %>% select(3, 5:15)

#select the rows required
rs1<- rs1 %>% slice(2, 4:11)
```

Finally, I converted the data it into a 'long' format for use with ggplot2.

```{r}
#convert to a 'long' data set
rs1_long <- rs1 %>% 
  gather(Year, Number, -Region)
```

The tibble below shows a short section of the data. 

```{r}
#shows the first 16 rows of the dataframe
head(rs1_long, 16)
```

I then prepared the data for the second visualisation; a pie chart of the number of rough sleepers in each region proportional to population, in 2019. This required importing a further data set that contained government data on the population of each region in England. This could be used to calculate a proprtoional value of rough sleepers.  

```{r, message=FALSE}
#import the dataset relating to number of rough sleepers for each region
rawdata <- read_excel("~/myfinalproject/processed/rs_statistics.xlsx")
#import the dataset relating to the population of eeach region
rawdata2<-read_excel("~/myfinalproject/processed/population.xlsx") 

#for the first data set pertaining to the number of rough sleepers
#select rows of interest
rs2<- rawdata %>% slice(2, 4:11)
#select the columns of interest
rs2<- rs2 %>% select (3, 14)

#convert to a 'long' format
rs2_long <- rs2 %>% 
  gather(Year, Number, -Region)

#for the data set containing the population for each region
#the rows and columns of interest are selected
rs3<- rawdata2 %>% slice(9, 23, 67, 92, 138, 173, 224, 258, 329)
rs3<- rs3 %>% select(2, 5)

#columns are renamed appropriately
rs3<- rs3 %>% 
  rename(
    Region = ...2,
   "2019" = ...5)

#convert the data to 'long' format
rs3_long <- rs3 %>% 
  gather(Year, Number, -Region)
#the Number column is converted to a numeric variable
rs3_long$Number <- as.numeric(as.character(rs3_long$Number))

#dataset is reordered to match the other data set
rs3_long<- rs3_long %>% arrange(Region)

#column of interest is then selected and renamed appropriately
rs3_long<- rs3_long %>% select(3)
rs3_long<- rs3_long %>%
    rename(Population = Number)

#then a new dataset is generated so the number of RS in each region proportional to population can be generated in the pie chart
#the two datasets are combined into a new table
newtable<- cbind(rs2_long, rs3_long)
newtable <- transform(newtable, #calculates % of the RS withi the region population
                      Percentage = Number / Population) 

#select the columns to be used for the pie chart visualisation
newtable <- newtable %>% select(1,2,5)

#first six lines of the new data set
head(newtable)
```

<span style="text-decoration:underline">Data Visualisation</span> 

Graph 1: Number of Rough Sleepers in England by Region (2010 - 2020)

```{r = graph1, message=FALSE, results='hide'}
#generate a colour palette with more than 8 colours by extending ggplot palette
nb.cols <- 9
mycolors <- colorRampPalette(brewer.pal(8, "Set2"))(nb.cols) 
  
#generates a line graph with several additional features
rs1plot <-ggplot(rs1_long, aes(x = Year, y = Number, color = Region, group = Region)) +
  geom_point() +
  geom_line(position = position_dodge(0.4)) + #clustered data points are distributed
  scale_fill_manual(values = mycolors) + #a colour scheme is added
  theme(axis.text.x = element_text(face= "bold", color="black", 
                                 size=7),
        axis.text.y = element_text(face= "bold", color="black", 
                                  size=7),
        axis.line = element_line(colour = "black", 
                    size = 1, linetype = "solid")) + #aesthetic of the x/y axis
  labs(title="Number of Rough Sleepers in England by Region (2010 - 2020)",
       x="Year", y = "Number of Rough Sleepers") + #labels the x/y axis and title for plot
  scale_y_continuous(breaks=c(0, 300, 600, 900, 1200, 1400), limits=c(0, 1400)) #adjust the scale

ggsave("figures/rs1plot.png") #saves the graph as an image in figs folder
```

```{r echo=FALSE, warning=FALSE}
rs1plot #generates the plot
```

Graph 2: Number of Rough Sleepers by Region proportional to population (2019) 

``` {r = second visualisation, message=FALSE, results='hide'}
#look at the data prepared previously
view(newtable)

#generate a dataframe of the data
rspie<- data.frame(newtable)

#generate a colour palette with more than 8 colours by extending ggplot palette
nb.cols <- 9 
mycolors <- colorRampPalette(brewer.pal(8, "Set2"))(nb.cols) 

#simple barplot is generated of the dataframe first
bp<- ggplot(rspie, aes(x="", y=Percentage, fill=Region))+
  geom_bar(width = 1, stat = "identity") + 
  geom_text(aes(label = paste(round(Percentage / sum(Percentage) * 100, 1), "%"), x = 1.7),
         position = position_stack(vjust = 0.5)) +
  ggtitle("help") +
  theme(axis.text = element_blank(),
        axis.title = element_blank(),
        axis.ticks = element_blank(),
        panel.grid  = element_blank(),
        legend.position="none",
        line = element_blank()) + #blank template theme is added
  scale_fill_manual(values = mycolors) #colour scheme is added matching plot above


#barplot is converted into a piechart
rspiechart <- bp + coord_polar(theta = "y", start=0) +
  theme_void() +
  ggtitle("Percentage of Rough Sleepers in England by Region proportional to population (2019)")

#saves the graph as an image in figs folder
ggsave("figures/rspiechart.png")
```

```{r echo=FALSE, warning=FALSE}
#generates the pie chart
rspiechart
```

<span style="text-decoration:underline">Summary</span>

The main visualisation show that the number of rough sleepers in England appears to have increased within all regions in England from 2010 to 2020. The supporting visualisation indicates that the largest of rough sleepers in England appear to be in London, and the South of England when taking into account the population as well as number of rough sleepers in each region. 
