---
title: "Data Visualisation Project"
author: "Seline Uran"
output: html_document
---
```{r = diagram1, echo=FALSE, message=FALSE}
#To begin load the following libraries and transfer the .zip file to the 'Documents' folder 
library(ggplot2)
library(readxl)
library(RColorBrewer)
library(scales)
library(dplyr)
library(tidyr)
library(tidyverse)
library(gridExtra)
library(RColorBrewer)
```

<span style="text-decoration:underline">Research Questions</span>

These graphs present government survey data on the number of rough sleepers in England to consider the following questions; 

Has the number of rough sleepers in England increased over the past decade? 

Is the number of sleepers observed equally across England in one year? 

<span style="text-decoration:underline">Data Origins</span>

The raw data, the number of rough sleepers in England from 2010 - 2020, was retrieved from https://www.gov.uk/government/statistics/rough-sleeping-snapshot-in-england-autumn-2020. This data is gathered by individuals in each local authority conducting either (1) a counts-based estimate (counting visible rough sleepers), (2) an evidence-based estimate that involves meeting with local community partners or (3) an evidence-based estimate that performs counts in a specific area. The measure is carried out on a single night between October 1st and November 30th each year. These measures do not provide an index of homelessness but represents an estimate of individuals sleeping rough (therefore do not include individuals who are 'sofa surfing', in temporary accommodation or in a homeless shelter/hostel) on a single night.

<span style="text-decoration:underline">Data Preparation</span>

After having removed all sheets from the excel document except for the sheet that contained the total number of rough sleepers I began data preparation. 

```{r}
#importing the raw data
rawdata <- read_excel("rs_statistics.xlsx")
```

The raw data detailed the number of rough sleepers within England and individual counts for each region of England and each Local Authority within each region. 

My aim was to visualise the number of rough sleepers in each region of England across the last 10 years. I selected these rows from the data set. Since I aimed to visualize the number of rough sleepers in each region individually I removed the rows containing summary counts for 'England', 'London' and the 'Rest of England'. Then I removed data associated with each Local Authority and the ONS codes for each Region and Local Authority. 

```{r}
#select the columns required
rs1<- rawdata %>% select(3, 5:15)

#select the rows required
rs1<- rs1 %>% slice(2, 4:11)
```

Finally, I converted the data it into a 'long' format for use with ggplot2.

```{r}
#convert to a 'long' data set
rs1_long <- rs1 %>% 
  gather(Year, Number, -Region)
```

The tibble below shows a short section of the data. 

```{r}
#shows the first 16 rows of the dataframe
head(rs1_long, 16)
```

I then prepared the data for the second visualisation. My aim was to create a pie chart of the number of rough sleepers in England by region.  

```{r, message=FALSE}
#select rows of interest
rs2<- rawdata %>% slice(2, 4:11)
#select the columns of interest
rs2<- rs2 %>% select (3, 14)

#convert to a 'long' format so that the data can be input to be visualised by ggplot
rs2_long <- rs2 %>% 
  gather(Year, Number, -Region)

#displays the first six rows of the tibble
head(rs2_long)
```

<span style="text-decoration:underline">Data Visualisation</span> 

Graph 1: Number of Rough Sleepers in England by Region (2010 - 2020)

```{r = graph1, message=FALSE, results='hide'}
#generate a colour palette with more than 8 colours by extending ggplot palette
nb.cols <- 9
mycolors <- colorRampPalette(brewer.pal(8, "Set2"))(nb.cols) 
  
#generates a line graph with several additional features
rs1plot <-ggplot(rs1_long, aes(x = Year, y = Number, color = Region, group = Region)) +
  geom_point() +
  geom_line(position = position_dodge(0.4)) + #clustered data points are distributed
  scale_fill_manual(values = mycolors) + #a colour scheme is added
  theme(axis.text.x = element_text(face= "bold", color="black", 
                                 size=7),
        axis.text.y = element_text(face= "bold", color="black", 
                                  size=7),
        axis.line = element_line(colour = "black", 
                    size = 1, linetype = "solid")) + #aesthetic of the x/y axis
  labs(title="Number of Rough Sleepers in England by Region (2010 - 2020)",
       x="Year", y = "Number of Rough Sleepers") + #labels the x/y axis and title for plot
  scale_y_continuous(breaks=c(0, 300, 600, 900, 1200, 1400), limits=c(0, 1400)) #adjust the scale

ggsave("figures/rs1plot.png") #saves the graph as an image in figs folder
```

```{r echo=FALSE, warning=FALSE}
rs1plot #generates the plot
```

Graph 2: Percentage of Rough Sleepers in England by Region (2019) 

``` {r = second visualisation, message=FALSE, results='hide'}
#generate a dataframe of the data
rspie<- data.frame(rs2_long)

#generate a colour palette with more than 8 colours by extending ggplot palette
nb.cols <- 9 
mycolors <- colorRampPalette(brewer.pal(8, "Set2"))(nb.cols) 

#simple barplot is generated of the dataframe first
bp<- ggplot(rspie, aes(x="", y=Number, fill=Region))+
  geom_bar(width = 1, stat = "identity") + 
  geom_text(aes(label = paste(round(Number / sum(Number) * 100, 1), "%"), x = 1.7),
         position = position_stack(vjust = 0.5)) +
  ggtitle("help") +
  theme(axis.text = element_blank(),
        axis.title = element_blank(),
        axis.ticks = element_blank(),
        panel.grid  = element_blank(),
        legend.position="none",
        line = element_blank()) + #blank template theme is added
  scale_fill_manual(values = mycolors) #colour scheme is added matching plot above


#barplot is converted into a piechart
rspiechart <- bp + coord_polar(theta = "y", start=0) +
  theme_void() +
  ggtitle("Percentage of Rough Sleepers in England by Region (2019)")

#saves the graph as an image in figs folder
ggsave("figures/rspiechart.png")
```

```{r echo=FALSE, warning=FALSE}
#generates the pie chart
rspiechart
```

<span style="text-decoration:underline">Summary</span>

The main visualisation show that the number of rough sleepers in England seems to have increased overall across all regions in England from 2010 to 2020 though there was a reduction in the number between 2018 and 2020. However, the data from 2019 on wards should be interpreted with caution. The counts conducted in 2019 were obtained during the COVID-19 pandemic following the government's 'Everyone In' campaign that aimed to temporarily house rough sleepers. As such, the reduction observed between 2019 - 2020 is likely a result of this policy. Despite this action the data indicates that there is still a large increase in number of rough sleepers in 2020 compared to 2010. The supporting visualisation indicates that the largest proportion of rough sleepers in England appear to be in London and the South of England (South East and South West). 


<span style="text-decoration:underline">Future Directions</span>

A recommendation for improvement on the current visualisation of the data set might be to consider the number of rough sleepers in each region proportional to the population of that region. This would allow for a more accurate comparison across regions of the number of rough sleepers. In addition, the age distribution and composition of gender and ethnicity of rough sleepers in each region were not visualised. These factors might reveal greater insight into the trends within the population of rough sleepers across each region in England and therefore those groups most at risk. 
